---
title: "Unit 2 Project"
author: "Leila Haidar"
date: "2026-02-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## **Lecture 6**

install.packages("tidyverse")
install.packages("palmerpenguins")

```{r}
library("tidyverse")
library("palmerpenguins")
penguins <- penguins %>%
  drop_na(flipper_length_mm, body_mass_g, bill_length_mm, bill_depth_mm, sex)
```

### **Q1**

```{r}
ggplot(penguins, aes(x = island)) +
  geom_bar(fill = "steelblue") +
  labs(x = "Island", y = "Count")
```

Biscoe has the most penguins

```{r}
ggplot(penguins, aes(x = island)) +
  geom_bar(fill = "steelblue") +
  labs(x = "Island", y = "Count") +
  theme_minimal(base_size = 16)
```

Adding theme_minimal() makes the graph look cleaner/more visually sound.

### **Q2**

```{r}
ggplot(penguins, aes(x = body_mass_g)) +
  geom_histogram(binwidth = 60,
                 fill = "mediumpurple3",
                 color = "mediumpurple1") +
  labs(x = "body mass (g)",
       y = "Count") +
  theme_minimal(base_size = 15)
```

The general shape of the distribution is normal with a right skew. Changing bin number changes the range of the x-axis, or body mass.

### **Q3**

```{r}
ggplot(penguins,
       aes(x = species,
           y = flipper_length_mm,
           color = species)) +
  geom_boxplot() +
  labs(x = "Species", y="Flipper Length (mm)")+
  theme_minimal()+
  scale_color_brewer(palette = "Paired")+
  theme(legend.position = "none")
```

Gentoo have longest flippers (~216 mm average)

```{r}
ggplot(penguins,
       aes(x = sex,
           y = body_mass_g,
           color = sex)) +
  geom_boxplot() +
  labs(x = "Sex", y="Body Mass (g)")+
  theme_minimal()+
  scale_color_brewer(palette = "Accent")+
  theme(legend.position = "none")
```

Males have on average a higher body mass.

```{r}
ggplot(penguins,
       aes(x = species,
           y = flipper_length_mm,
           fill = species)) +
  geom_boxplot() +
  labs(x = "Species", y="Flipper Length (mm)")+
  theme_minimal()+
  theme(legend.position = "none")
```

### **Q4**

```{r}
ggplot(penguins,
       aes(x = bill_length_mm,
           y = bill_depth_mm)) +
  geom_point(size = 2)+
  geom_point(color = "mediumpurple1")+
  labs(x = "Bill Length (mm)",
       y = "Bill Depth (mm)")+
    theme_minimal()
```

The relationship of the scatterplot is unclear. Adding a color makes the scatterplot slightly more readable, however overall still unclear.

```{r}
ggplot(penguins,
       aes(x = bill_length_mm,
           y = bill_depth_mm)) +
  geom_point(size = 2)+
  geom_point(color = "mediumpurple1")+
  labs(x = "Bill Length (mm)",
       y = "Bill Depth (mm)")+
  theme_minimal()+
  geom_smooth(method="lm", na.rm=TRUE)
```

Adding the line of best fit shows a slight negative relationship between Bill Length and Depth. Overall, however, the line does not fit the data well.

### **Q5**

```{r}
ggplot(penguins,
       aes(x = flipper_length_mm,
           y = body_mass_g,
           color = species)) +
  geom_point(size = 1)+
  facet_wrap(~species)+
  labs(x = "Flipper Length (mm)",
       y = "Body Mass (g)")+
  theme_minimal()+
  scale_color_brewer(palette="Set2")
```

The Gentoo species shows the strongest relationship with Body mass and Flipper length.

```{r}
ggplot(penguins,
       aes(x = flipper_length_mm,
           y = body_mass_g,
           color = species)) +
  geom_point(size = 1)+
  facet_wrap(sex~species)+
  labs(x = "Flipper Length (mm)",
       y = "Body Mass (g)")+
  theme_minimal()+
  scale_color_brewer(palette="Set2")
```

The pattern indicates that male Gentoo penguins have the highest body mass. Additionally, body mass is generally higher among males, and Gentoo penguins are larger for both sexes.

```{r}
ggplot(penguins,
       aes(x = flipper_length_mm,
           y = body_mass_g,
           color = species)) +
  geom_point(size = 1)+
  facet_wrap(~species)+
  facet_wrap(~sex)+
  labs(x = "Flipper Length (mm)",
       y = "Body Mass (g)")+
  theme_minimal()+
  scale_color_brewer(palette="Set2")
```

### **Q6**

```{r}
files<-ggplot(penguins,
       aes(x = flipper_length_mm,
           y = body_mass_g,
           color = island)) +
  geom_point(size = 1)+
  facet_wrap(sex~island)+
  labs(x = "Flipper Length (mm)",
       y = "Body Mass (g)")+
  theme_dark()+
  scale_color_brewer(palette="Blues")
files
```

### **Q7**

- The most fun part of the exercise was being able to customize the colors of different graphs, and create my own graph.
- The most confusing part of the exercise was determining the parameters format for each type of graph.

## **Lecture 7**

```{r}
library("palmerpenguins")
library("dplyr")
library("tidyverse")
library(readr)
penguins_messy <- read_csv("penguins_messy.csv")
view(penguins_messy)
```

### **Q1**

```{r}
str_detect(penguins_messy$species_messy, "adelie")
?str_detect
```

Detects matching values, or if any values contain "adelie"

```{r}
str_replace_all(penguins_messy$sex_messy, "\\?", "")
?str_replace_all
```

Replaces matching penguin sex values with a new text

```{r}
str_squish(penguins_messy$island_messy)
?str_squish
```

Removes whitespace, or spaces/line breaks, in the island category.

### **Q2**

```{r}
penguins_messy<-penguins_messy %>%
  mutate(valid_id = str_detect(sample_id, "PAL-[0-9]{3}"))

penguins_messy %>%
  pull(valid_id) %>%
  sum
```

### **Q3**

```{r}
penguins_messy<- penguins_messy %>%
  filter(valid_id)
penguins_messy
```

### **Q4**

```{r}
penguins_messy<-penguins_messy %>%
  mutate(
    file_date = str_extract(file_name, "^[0-9]{4}-[0-9]{2}-[0-9]{2}"),
    file_id = str_extract(sample_id, "PAL-[0-9]{3}"))

penguins_messy %>%
  filter(row_number() <= 10) %>%
  pull(file_date)
penguins_messy %>%
  filter(row_number() <= 10) %>%
  pull(file_id)
```

### **Q5**

```{r}
penguins_messy<-penguins_messy %>%
  mutate(island=
  str_squish(island_messy))

unique(penguins_messy$island)
```

### **Q6**

```{r}
penguins_messy<-penguins_messy %>%
  mutate(sex=
  str_replace_all(sex_messy,"\\?","") %>% str_to_upper)
```

*OR*

```{r}
penguins_messy<-penguins_messy %>%
  mutate(sex=
  str_replace_all(sex_messy,"\\?",""))%>%
  mutate(sex=
  str_to_upper(sex))

unique(penguins_messy$sex)
```

### **Q7**

```{r}
penguins_messy<-penguins_messy %>%
  mutate(bill_mm=
  str_replace_all(notes,"\\.*bill=|mm",""))

unique(penguins_messy$bill_mm)
```

### **Q8**

*Most fun/interesting part of exercise? Most confusing or difficult?*

- The most interesting part was finally figuring out how to use the different str functions in the mutate function.

- The most confusing/difficult part was also figuring out how to use the different str functions in the mutate function.

*Do you feel you have a general idea of how regular expressions are used?*

- Yes, though I will likely practice more because I only started to understand it towards the last question.

## **Lecture 8**

```{r}
penguins <- palmerpenguins::penguins
library("tidyverse")
library("palmerpenguins")
```

### **Q1**

```{r}
penguins_clean <- penguins %>%
  drop_na(bill_length_mm,sex)

nrow(penguins_clean)
nrow(penguins)
```

333 rows (oenguins clean) vs 344 (penguins)

### **Q2**

```{r}
penguins_dup <- penguins_clean %>%
  bind_rows(slice_sample(penguins_clean, n = 30, replace = TRUE))
nrow(penguins_dup)
penguins_dup %>% distinct %>% nrow
```

### **Q3**

```{r}
penguins_new <- penguins_clean %>%
  select(species, island, sex, where(is.numeric))
length(penguins_new)
```

8 total collumns, with 333 rows (penguins_new)

### **Q4**

```{r}
penguins_bill <- penguins_clean %>%
  select(starts_with("bill"),ends_with("_mm"))
penguins_bill

penguins_bill <- penguins_clean %>%
  select(matches("bi.*|fli.*"))
penguins_bill
```

### **Q5**

```{r}
penguins_species <- penguins_clean %>%
  drop_na %>%
  group_by(species) %>%
  summarize(across(where(is.numeric),mean))
penguins_species
```

The code across(where(is.numeric), …) is better than typics each column by hand because it is more efficient/faster.

### **Q6**

```{r}
penguins_clean<-penguins_clean %>%
  mutate(bill_length=
           case_when(
             bill_length_mm<40~"short",
             bill_length_mm<50~"medium",
             bill_length_mm>=50~"long"))
penguins_clean
penguins_clean$bill_length
```

### **Q7**

```{r}
instrument_meta <- tibble(
  instrument_id = c("caliper_A", "caliper_B", "caliper_C"),
  calibration_mm = c(0.2, 0.5, 1.0),
  manufacturer = c("Mitutoyo", "Fowler", "Generic")
)
instrument_meta

penguins_instrumented<-penguins_clean %>%
  drop_na(bill_length_mm)%>%
  mutate(instrument_id=
         case_when(
           year==2007~"caliper_A",
           year==2008~"caliper_B",
           year==2009~"caliper_C"))
penguins_instrumented

penguins_instrumented<-penguins_instrumented %>%
  left_join(instrument_meta,by="instrument_id")

penguins_instrumented
length(penguins_instrumented)
length(penguins_clean)
```

penguins_instrumented is 12 rows and penguins_clean is 9 rows.

### **Q8**

- Most fun/interesting part of the exercise: I liked making different dataframes and testing the different functions from lecture.
- Most confusing/difficult part: The second half of number 7, and figuring out how to get left_join to work.
- I feel like these general "survival tips" will be useful in future data analyses because they can allow for needed edits in a quick, efficient way.
